Wealth dynamics in the presence of\\network structure and primitive cooperation

### 2019-07-23 manuscript/bodycssa19.tex
minor changes, Will's comments addressed

\section{Introduction}
In recent years, the concerns and debates regarding wealth inequality and socioeconomic mobility have been one of the few unifying issues dominating the extremely polarized public spheres of the Global North. While economic and political inequality used to be discussed in heterodox economics circles, the contentious discussions on the topic within mainstream economics since the publication of Piketty's book~\cite{piketty2017capital} suggest a lack of consensus about basic foundational questions like the origins and persistence of economic inequality. Not surprisingly, traditional theories and tools of macro and micro economics are now being diagnosed for their limitations. Simultaneously, insights from related disciplines, along with novel models for scientific inquiry not usually associated with traditional econometrics, are being taken more seriously. The work presented here shares this spirit by integrating substantive ideas from anthropology, economic sociology~\cite{granovetter2017society} and urban sociology~\cite{sampson2012great} and using modeling approaches from computational social science~\cite{hedstrom2018}, computational economics~\cite{tesfatsion} and analytical sociology~\cite{hedstrom2011oxford} to explore the origins of inequality in a simple model of wealth dynamics in the presence of social structures.

The current work originated in our attempts to incorporate and tease out the effects of social structures in simple models of wealth dynamics in the presence of environmental stochasticity, and a simple form of resource pooling. Specifically, we are interested in studying how structure and dynamics interact to produce and sustain inequality and to influence robustness to scarcity shocks. We answer the former question using Gini coefficients and the latter using survival time type analysis. As we discuss in section 3, while Gini was found to be of limited value, survival time analysis produced more insights on both questions of inequality and robustness.

@@

Models constructed here, like those used in scientific inquiry in general, serve a specific purpose. In this work, we construct simple toy models to reproduce certain aspects of non-trivial wealth inequality distributions in the presence and absence of primitive forms of cooperation, clearly delineating the role of network structure in generating wealth inequality. We make no suggestions that these models \textit{explain} the phenomena of interest; we are only interested in constructing the simplest possible models, with no detailed empirical grounding, but with the potential to generate heterogeneous wealth outcomes reminiscent of ones obtained using more sophisticated models. In doing so, we aspire to shed light on the true social and economic mechanisms underlying the genesis and persistence of economic inequality, and network structure alone can produce large differences in economic outcomes.

@@

For such ensemble of particles, the primary driver of difference in paths (life histories) and wealth is the Brownian noise intensity $D$; greater the $D$ leads greater the diversity (and hence Gini). However, Gini is a measure that is dependent on absolute magnitude. So, if the drift ($v$) is large, then the variation generated by
$D$ gets washed out by $v$\footnote{Preliminary investigations suggest that for simple non-network ensembles, Gini either stays close to $0$ or $1$. We suspect that this is because of the constant wealth growth rate used in our models, Gini is a partially useful measure. This is unlike in macroeconomic models where exponential growth rate gives rise an appearance of non-trivial Gini coefficients}.  On the other hand, the mathematical form of expressions for $\mathcal{S}(t)$ suggest a clear dependence on $D$ which separates the population of particles that are not in a coalescent and the population of particles that are in one. Particles reaches the absorbing state (death of the agent) without being part of a coalescent are called isolates; the non-isolates, the particles that are in a coalescent when they reach the absorbing state are called protos.

@@

To break the homogeneity of this particle ensemble, we propose to let the particles interact with only a random subset of other populations. In other words, we embed the particles on a network (graph); particles interact (form protos) with only their neighbors. In this work, we focus on the simplest \textit{textbook} example of network: Erd\H{o}s-R\'{e}nyi network model~\cite{newman2018networks}. ER models have only one parameter ($\lambda$), which is the \textit{average expected degree} of any given node. This determines the density of connections in the network, and in this setting determines the expected number of particles available for a particle to coalesce with and form a proto. Commonsense reasoning suggests that a larger $\lambda$ parameter may help increase the average lifetimes of individual agents by enabling greater likelihood of contact with other particles.

@@

\begin{enumerate}
\itemsep.1em
\item Create $N$ agents, each with a random initial wealth (uniformly distributed from 0 to \texttt{init\_max}), and related to one another as per a random ER network with parameter $\lambda$.
\item \textbf{Stages 1 and 2:} Repeat until all non-isolate agents\footnote{An \textbf{isolate} agent is one with no graph neighbors.} are members of a proto:
    \begin{enumerate}
    \itemsep.1em
    \item Each agent $A$ whose current wealth $\ge x_{\textrm{thresh}}$, and who is not currently a member of a proto, chooses at random one of its graph neighbors (call it $B$) whose wealth \textit{also} exceeds $x_{\textrm{thresh}}$. (If there are no such neighbors, proceed to the next agent.) If $B$ is already in a proto, have $A$ join $B$'s proto. If not, have $A$ and $B$ form a new proto.
    \item Each agent gains an amount of wealth equal to $(\texttt{salary} - m + \epsilon)$, where $\epsilon \sim \mathcal{N}(0,\sigma^2)$ (white noise).\footnote{Note that this ``gain'' could be negative, in which case the agent, and possibly its proto, may be subject to death exactly as in step (3).}
    \item Each agent \textit{that is in a proto} donates all its wealth in excess of $x_{\textrm{thresh}}$ to that proto's balance. (Agents not in a proto maintain their current wealth.)
    \end{enumerate}
    \item \textbf{Stage 3 (starvation):} Repeat until all agents are dead:
        \begin{enumerate}
        \itemsep.1em
        \item Each agent \textit{loses} an amount of wealth equal to $(m + \epsilon), \epsilon \sim \mathcal{N}(0,\sigma^2)$.
        \item If an agent's wealth would drop below zero as a result of this loss, and if it is not a member of a proto, it dies and is removed from the simulation. If it \textit{is} a member of a proto, it withdraws the necessary amount from its proto's balance to remain at zero wealth. If the proto does not have sufficient funds to cover the loss, both the agent and the proto die and are removed from the simulation.
        \end{enumerate}
\end{enumerate}

@@

>>>>

\section{Introduction}
In recent years, the concerns and debates regarding wealth inequality and socioeconomic mobility have been one of the few unifying issues dominating the extremely polarized public spheres of the Global North. While economic and political inequality used to be discussed in heterodox economics circles, the contentious discussions on the topic within mainstream economics since the publication of Piketty's book~\cite{piketty2017capital} suggest a lack of consensus about basic foundational questions like the origins and persistence of economic inequality. Not surprisingly, traditional theories and tools of macro and micro economics are now being diagnosed for their limitations. Simultaneously, insights from related disciplines, along with novel models for scientific inquiry not usually associated with traditional econometrics, are being taken more seriously. This work contributes to this effort by integrating substantive ideas from anthropology, economic sociology~\cite{granovetter2017society} and urban sociology~\cite{sampson2012great} and using modeling approaches from computational social science~\cite{hedstrom2018}, computational economics~\cite{tesfatsion} and analytical sociology~\cite{hedstrom2011oxford} to explore the origins of inequality in a simple model of wealth dynamics, with social structures and a primitive form of cooperation .

The current work originated in our attempts to incorporate and tease out the effects of social structures in simple models of wealth dynamics in the presence of environmental stochasticity and a simple form of resource pooling. We are interested in the interaction between structure and dynamics, their role in the production and persistence of inequality, and how they influence the system's robustness to scarcity shocks. We answer the former question using Gini coefficients and the latter using survival time type analysis. As we discuss in section 3, while Gini was found to be of limited value, survival time analysis produced more insights on both questions of inequality and robustness.

@@

Models constructed here, like those used in scientific inquiry in general, serve a specific purpose. In this work, we construct simple toy models to reproduce certain aspects of non-trivial wealth inequality distributions in the presence and absence of primitive forms of cooperation, clearly delineating the role of network structure in generating wealth inequality. We make no suggestions that these models \textit{explain} the phenomena of interest; we are only interested in constructing the simplest possible models, with no detailed empirical grounding, but with the potential to generate heterogeneous wealth outcomes reminiscent of ones obtained using more sophisticated models. As we discuss later, these simple mechanisms are limited in their ability to stabilize wealth and inequality. Despite this, the model still sheds light on the true social and economic mechanisms underlying the genesis and persistence of economic inequality, and network structure alone can produce large differences in economic outcomes.

@@

For such ensemble of particles, the primary driver of difference in paths (life histories) and wealth is the Brownian noise intensity $D$; greater the $D$ leads greater the diversity (and hence Gini). However, Gini is a measure that is dependent on absolute magnitude. So, if the drift ($v$) is large, then the variation generated by
$D$ gets washed out by $v$\footnote{Preliminary investigations suggest that for simple non-network ensembles, Gini either stays close to $0$ or $1$. We suspect that this is because of the constant wealth growth rate used in our models, Gini is a partially useful measure. This is unlike in macroeconomic models where exponential growth rate gives rise to stable non-trivial Gini coefficients}. On the other hand, the mathematical form of expressions for $\mathcal{S}(t)$ suggest a clear dependence on $D$ which separates the population of particles that are not in a coalescent and the population of particles that are in one. Particles reaching the absorbing state (death of the agent) without being part of a coalescent are called isolates\footnote{In the context of a system with non-trivial network structure, an \textbf{isolate} agent is one with no graph neighbors.}; the non-isolates, the particles that are in a coalescent when they reach the absorbing state are called protos.

@@

To break the homogeneity of this particle ensemble, we propose to let the particles interact with only a random subset of other populations. In other words, we embed the particles on a network (graph); particles interact (form protos) with only their neighbors. In this work, we focus on the simplest \textit{textbook} example of network: Erd\H{o}s-R\'{e}nyi network model~\cite{newman2018networks}. ER models have only one parameter ($\lambda$), which is the \textit{average expected degree} of any given node. This determines the density of connections in the network, and in this setting determines the expected number of particles available for a particle to coalesce with and form a proto. Commonsense reasoning suggests that a larger $\lambda$ parameter may help increase the average lifetimes of individual agents by enabling proto formation by increasing the likelihood of contact with other particles.

@@

\begin{enumerate}
\itemsep.1em
\item Create $N$ agents, each with a random initial wealth (uniformly distributed from 0 to \texttt{init\_max}), and related to one another as per a random ER network with parameter $\lambda$.
\item \textbf{Stages 1 and 2:} Repeat until all non-isolate agents are members of a proto:
    \begin{enumerate}
    \itemsep.1em
    \item Each agent $A$ whose current wealth $\ge x_{\textrm{thresh}}$, and who is not currently a member of a proto, chooses at random one of its graph neighbors (call it $B$) whose wealth \textit{also} exceeds $x_{\textrm{thresh}}$. (If there are no such neighbors, proceed to the next agent.) If $B$ is already in a proto, have $A$ join $B$'s proto. If not, have $A$ and $B$ form a new proto.
    \item Each agent gains an amount of wealth equal to $(\texttt{salary} - m + \epsilon)$, where $\epsilon \sim \mathcal{N}(0,\sigma^2)$ (white noise).\footnote{Note that this ``gain'' could be negative, in which case the agent, and possibly its proto, may be subject to death exactly as in step (3).}
    \item Each agent \textit{that is in a proto} donates all its wealth in excess of $x_{\textrm{thresh}}$ to that proto's balance. (Agents not in a proto maintain their current wealth.)
    \end{enumerate}
    \item \textbf{Stage 3 (starvation):} Repeat until all agents are dead:
        \begin{enumerate}
        \itemsep.1em
        \item Each agent \textit{loses} an amount of wealth equal to $(m + \epsilon), \epsilon \sim \mathcal{N}(0,\sigma^2)$.
        \item If an agent's wealth would drop below zero as a result of this loss, and if it is not a member of a proto, it dies and is removed from the simulation. If it \textit{is} a member of a proto, it withdraws the necessary amount from its proto's balance to remain at zero wealth. If the proto does not have sufficient funds to cover the loss, both the agent and the proto die and are removed from the simulation.
        \end{enumerate}
\end{enumerate}

@@

An obvious and important limitation of the current model is the lack of stability in wealth inequality in the population; neither the ER graph structure nor the Brownian noise stabilizes procured wealth. Therefore, Additional mechanisms are required to maintain the inequality.

### 2019-07-23 manuscript/specnet_conffmt.tex
(Remove weird title line break.)
minor changes, Will's comments addressed

\begin{document}
\title{Wealth dynamics in the presence of\\network structure and primitive cooperation}
%\titlenote{Produces the permission block, and
 % copyright information}
%\subtitle{Extended Abstract}
%\subtitlenote{The full version of the author's guide is available as
 % \texttt{acmart.pdf} document}

@@

\begin{abstract}
We study wealth accumulation dynamics in a population of heterogeneously mixed agents with a capacity for a certain primitive form of cooperation enabled by static network structures. Despite their simplicity, the stochastic dynamics generate inequalities in wealth reminiscent of real-world social systems even in a fully mixed population. A simple form of cooperation is introduced and is shown to enhance the viability of agents by embedding such dynamics in a network; the impact of social structures on the origins and persistence of inequality can be teased out easily. The models developed here complement traditional modeling approaches based on grid worlds.

>>>>

\begin{document}
\title[Wealth dynamics in the presence of network structure and primitive cooperation]{Wealth dynamics in the presence of\\network structure and primitive cooperation}
%\titlenote{Produces the permission block, and
 % copyright information}
%\subtitle{Extended Abstract}
%\subtitlenote{The full version of the author's guide is available as
 % \texttt{acmart.pdf} document}

@@

\begin{abstract}
We study wealth accumulation dynamics in a population of heterogeneously mixed agents with a capacity for a certain primitive form of cooperation enabled by static network structures. Despite their simplicity, the stochastic dynamics generate inequalities in wealth reminiscent of real-world social systems even in a population without social structure. A simple form of social structure and cooperation is introduced and is shown to enhance the viability of agents. The interaction between social structures and dynamics illuminates their role in the generation and persistence of inequality. The models developed here complement traditional modeling approaches based on grid worlds.
